---
output: github_document
editor_options: 
  markdown: 
    wrap: sentence
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# SimpleMediation

<!-- badges: start -->

<!-- badges: end -->

The goal of SimpleMediation is to implement simple linear mediation with bootstrapping to estimate the error.
It provides functions the run simple mediation with bootstrapping.

## Installation

You can install the development version of SimpleMediation from a file, later GIt will be used:

``` r
# install.packages("devtools")
devtools::install_github("timothym/SimpleMediation")
```

## Usage

Mediation is examining the impact of a variable (M) on the relationship between two others (X) and (Y).
This package provides a function `simple_mediation()` to do this and give an estimate for the uncertainty in these relationships.


```{r}
x <- c(1,2,3,4,5,6,7)
m <- c(1,2.3,4.2,4.9,5.6,6,6.7)
y <- c(3.4,5.4,6.5,7.9,8,10,11)
df <- data.frame(x,m,y)
simple_mediation(df)
```

The returned values are the estimate of the total, direct, and indirect effects along with their bias and standard error.
