---
title: "Mediation_Details"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mediation_Details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SimpleMediation)
```

This package is code for my thesis on mediation in a time series context. This vignette details the theory behind mediation and the functions that I am writing for the thesis. This vignette will help link the theory with the specific functions that I am writing.

# Basic mediation functions

The functions `run_mediation`, and `simple_mediation` are both for normal mediation on continuous variables. `run_mediation` runs a simple mediation on three variables, and `simple_mediation` uses it to perform bootstrapping in a simple mediation context and gives an estimate for the variance and covariance of the mediation.

# Mediation for ARX models

We denote an ARX(1) model as:

$$
X_t = \phi X_{t-1} + \psi Z_{t-1} + \epsilon_t
$$

Where: $\epsilon_t \sim N(0,\sigma^2)$

We can examine the autocovariance of an ARX(1) model. If we use the above definition the ARX(1) model is: $X_t = \phi X_{t-1} + \psi Z_{t-1} + \epsilon_t$, and suppose that $X_0 \sim WN(0, \sigma^2_X(\frac{1}{1-\phi})+\sigma^2_Z(\frac{\psi^2}{1-\psi^2}))$ and we find the autocovariance:

$$
\begin{align}
\gamma (0) &= \sigma^2_X(\frac{1}{1-\phi})+\sigma^2_Z(\frac{\psi^2}{1-\psi^2}) \\
\gamma(1) &= Cov(X_{t+1},X_t) \\
&= Cov(\phi X_t + \psi Z_t + \epsilon_t, X_t) \\
&= \phi Cov(X_t, X_t) + \psi Cov(Z_t,X_t) + Cov(\epsilon_t, X_t) \\
&= \phi \gamma(0) + \psi Cov(Z_t,X_t)
\end{align}
$$

This calculation depends on the value of $Cov(Z_t,X_t)$:

$$
\begin{align}
Cov(Z_t,X_t) &= Cov(Z_t, \phi X_{t-1} + \psi Z_{t-1} + \epsilon_{z,t}) \\
&= \phi Cov(Z_t,X_{t-1}) + \psi Cov(Z_t,Z_{t-1}) + Cov(Z_t,\epsilon_{z,t}) \\
&= \phi Cov(Z_t, X_{t-1}) + \psi\gamma_z(1) \\
&= \phi(\phi Cov(Z_t,X_{t-2}) + \psi \gamma_z (2)) + \psi \gamma_z (1)
\end{align}
$$

## ARX in Mediation

For mediation we define $X$ to be AR(1) and $M$ and $Y$ to be ARX(1):

$$
\begin{align}
X_t &= \alpha X_{t-1} + \epsilon_{t,X}\\
M_t &= \beta_0 M_{t-1} +\beta_1 X_{t-1} +\epsilon_{t,M} \\
Y_t &= \gamma_0 Y_{t-1} + \gamma_1 M_{t-1} + \gamma_2 X_{t-1} + \epsilon_{t,Y}
\end{align}
$$

### Initial Values

For an AR(1) model the initial point $X_0$ follows a distribution with mean 0, and variance $\sigma^2/(1-\phi^2)$. In the mediation context we will need initial values for $X$, $M$, and $Y$.

Since $X$ is an AR(1) model, we define the initial value to be:

$$
X_0\sim \left( 0,\frac{\sigma^2}{(1-\phi^2)} \right)
$$

We also need an initial value for $M$, which is an ARX(1) process. The mean of the initial value is still 0, but the variance is now also dependent on $X$.

$$
\begin{align}
Var(M_t) &= Var(\beta_0 M_{t-1} + \beta_1 X_{t-1} + \epsilon_{t,M}) \\
&= Var( \epsilon_{t,M} + \beta_0 \epsilon_{t-1,M} + \beta_0^2 \epsilon_{t-2,M} + \cdots +
\beta_1 \epsilon_{t-1,X} + \beta_1^2 \epsilon_{t-2,X} + \cdots ) \\
&= \sigma^2_M (1 + \beta_0 + \beta_0^2+ \cdots) + \sigma^2_X (\beta_1 + \beta_1^2 + \cdots) \\
&= \sigma^2_M \left( \frac{1}{1-\beta_0^2} \right) + \sigma^2_X \left( \frac{\beta_1^2}{1-\beta_1^2} \right)
\end{align}
$$ 

So we get an initial value for $M$ of $M_0 \sim \left( 0, \left( \frac{1}{1-\beta_0^2} \right) + \sigma^2_X \left( \frac{\beta_1^2}{1-\beta_1^2} \right) \right)$.

Following a similar argument we get an initial value for $Y$ of $Y_0 = \left( 0 , \sigma^2_Y \left( \frac{1}{1-\gamma_0^2} \right) + \sigma^2_M \left( \frac{\gamma_1^2}{1-\gamma_1^2} \right) + \sigma^2_X \left( \frac{\gamma_2^2}{1-\gamma_2^2} \right) \right)$.

We note that the values of $Y$ are dependent on two previous values of $X$, and so we also need an initial value for $Y_1$. If we say that the initial value of $Y_0$ is $Y_0 = \left( 0 , \sigma^2_Y \left( \frac{1}{1-\gamma_0^2} \right) + \sigma^2_M \left( \frac{\gamma_1^2}{1-\gamma_1^2} \right) + \sigma^2_X \left( \frac{\gamma_2^2}{1-\gamma_2^2} \right) \right)$. Then the value of $Y_1$ is  $X_1 \sim \left( \gamma_0 (\sigma^2_Y \left( \frac{1}{1-\gamma_0^2} \right) + \sigma^2_M \left( \frac{\gamma_1^2}{1-\gamma_1^2} \right) + \sigma^2_X \left( \frac{\gamma_2^2}{1-\gamma_2^2} \right) ) , \sigma^2_Y \left( \frac{1}{1-\gamma_0^2} \right) + \sigma^2_M \left( \frac{\gamma_1^2}{1-\gamma_1^2} \right) + \sigma^2_X \left( \frac{\gamma_2^2}{1-\gamma_2^2} \right) \right)$

### Autocovariances

We cannot use the same method we did for simple mediation since substituting in the values for $X_{t-1}$ and $M_{t-1}$ gets us:

$$
\begin{align}
Y_t &= \gamma_0(\gamma_0 Y_{t-2} + \gamma_1 M_{t-2} + \gamma_2 X_{t-2} + \epsilon_{t-1,Y}) + \gamma_1(\beta_0 M_{t-2} +\beta_1 X_{t-2} +\epsilon_{t-1,M}) + \gamma_2 (\alpha X_{t-2} + \epsilon_{t-1,X}) + \epsilon_{t,Y} \\
&= \gamma_0^2 Y_{t-2} + (\gamma_0 + \beta_0) \gamma_1 M_{t-2} + (\gamma_0 \gamma_2 + \gamma_1 \beta_1 + \gamma_2 \alpha)X_{t-2} + (\gamma_0 \epsilon_{t-1,Y} + \gamma_1 \epsilon_{t-1,M} + \gamma_2 \epsilon_{t-1,X} + \epsilon_{t,Y})
\end{align}
$$

From which we note that there are implicit $X_{t-3}$'s in $Y_{t-2}$ and $M_{t-2}$. Although this does allow us to make a statement about how $Y_{t}$ is related to $X_{t-2}$. The direct effect of $X_{t-2}$ on $Y_t$ is $\gamma_0 \gamma_2 + \gamma_2 \alpha$, and the indirect effect of $X_{t-2}$ on $Y_t$ is $\gamma_1 \beta_1$.
