---
title: "Mediation_Details"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mediation_Details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SimpleMediation)
```

# Mediation for ARX models

## ARX in Mediation

For mediation we define $X$ to be AR(1) and $M$ and $Y$ to be ARX(1):

$$
\begin{align}
X_t &= \alpha X_{t-1} + \epsilon_{t,X}\\
M_t &= \beta_0 M_{t-1} +\beta_1 X_{t-1} +\epsilon_{t,M} \\
Y_t &= \eta_0 Y_{t-1} + \eta_1 M_{t-1} + \eta_2 X_{t-1} + \epsilon_{t,Y}
\end{align}
$$

In this context, since $X_t$, $M_t$, and $Y_t$ are all AR based models we also define $\alpha$, $\beta_0$, and $\eta_0$ to be bounded between -1 and 1:

$$
\alpha,\beta_0,\eta_0 \in (-1,1)
$$

It is reasonable for us to assume that each of these process is stationary. While many time series are not actually stationary, there are a wide variety of techniques that transform a non-stationary time series into a stationary one, it is reasonable for us to assume that when using this process the starting time series can be turned into stationary ones.

We also define $\epsilon$ from these models to be white noise with the following distribution:

$$
\epsilon_{t,X} \sim N(0,\sigma^2_X)\\
\epsilon_{t,M} \sim N(0,\sigma^2_M)\\
\epsilon_{t,Y} \sim N(0,\sigma^2_Y)
$$

Here we are also assuming that the errors are uncorrelated with each other, since they follow a white noise pattern. We also note that this implies $\epsilon_{t,X}$ is uncorrelated with $\epsilon_{t-1,X}$

### Initial Values

For an AR(1) model the initial point $X_0$ follows a distribution with mean 0, and variance $\sigma^2/(1-\alpha^2)$. In the mediation context we will need initial values for $X$, $M$, and $Y$.

Since $X$ is an AR(1) model, we define the initial value to be:

$$
X_0\sim N \left( 0,\frac{\sigma^2_X}{(1-\alpha^2)} \right)
$$

We also need an initial value for $M$, which is an ARX(1) process. The mean of the initial value is still 0, but the variance is now also dependent on $X$.

#### Variance of M

The variance of M can be calculated using the properties of variance as:

$$
\begin{align}
Var(M_t) &= Var(\beta_0 M_{t-1} + \beta_1X_{t-1} + \epsilon_{t,M}) \\
&= \beta_0^2 Var(M_{t-1}) + \beta_1^2 Var(X_{t-1}) + Var(\epsilon_{t,M}) + 2Cov(M_{t-1},X_{t-1}) + 2Cov(M_{t-1},\epsilon_{t,M}) + 2Cov(X_{t-1},\epsilon_{t,M})
\end{align}
$$

We now examine the covariances, and our assumptions of the models and their errors allow us to set some of them to 0.

-   $Cov(M_{t-1}, X_{t-1}) = 0$ since $M_{t-1}$ depends on $X_{t-2}$ and further back $X$'s, but not $X_{t-1}$, so this covariance is zero.
-   $Cov(M_{t-1},\epsilon_{t,M}) = 0$, since $M_{t-1}$ is a function of $\epsilon_{t-j,M}$ for $j \in [1,t]$, and from our assumptions $Cov(\epsilon_{t,M},\epsilon_{t-i,M}) = 0$ for $i \neq 0$.
-   $Cov(X_{t-1}, \epsilon_{t,M}) = 0$ Since $X_{t-1}$ only depends on $\epsilon_{t-i,X}$ for $i \in [1,t]$, that is we get a string of covariances between $\epsilon_{t,M}$ and $\epsilon_{t-i,X}$ which all have covariance 0 from our assumptions in setting up the system.

Using these results we continue the calculation of the variance of $M_t$:

$$
\begin{align}
Var(M_t) &= \beta_0^2 Var(M_{t-1}) + \beta_1^2 Var(X_{t-1}) + Var(\epsilon_{t,M}) + 2Cov(M_{t-1},X_{t-1}) + 2Cov(M_{t-1},\epsilon_{t,M}) + 2Cov(X_{t-1},\epsilon_{t,M}) \\
Var(M_t) &= \beta_0^2 Var(M_{t-1}) + \beta_1^2 Var(X_{t-1}) + Var(\epsilon_{t,M}) + 0 + 0 + 0 \\
\end{align}
$$

We use the variance of $X_{t-1}$ we calculated earlier, and make the assumption of constant variance: $Var(M_{t-i}) = Var(M_{t-j})$, for $i \neq j$. We have also previously defined $Var(\epsilon_{t,M}) = \sigma^2_M$.

$$
\begin{align}
Var(M_t) - \beta_0^2 Var(M_{t-1}) &= \beta_1^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X + \sigma^2_M \\
Var(M_t)(1-\beta_0^2) &= \beta_1^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X + \sigma^2_M \\
Var(M_t) &= \frac{\beta_1^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X + \sigma^2_M}{(1-\beta_0^2)}
\end{align}
$$

So we get an initial value for $M$ of $M_0 \sim N \left( 0, \frac{\beta_1^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X + \sigma^2_M}{(1-\beta_0^2)} \right)$.

#### Variance of Y

Following a similar argument we get an initial value for $Y$:

$$
\begin{align}
Var(Y_t) &= Var(\eta_0Y_{t-1} + \eta_1 M_{t-1} + \eta_2 X_{t-1} + \epsilon_{t,Y}) \\
&= \eta_0^2Var(Y_{t-1}) + \eta_1^2 Var(M_{t-1}) + \eta_2^2 Var(X_{t-1}) \\
&+ 2Cov(Y_{t-1}, M_{t-1}) + 2Cov(Y_{t-1},X_{t-1}) + 2Cov(Y_{t-1},\epsilon_{t,Y}) \\
&+ 2Cov(M_{t-1}, X_{t-1}) + 2Cov(M_{t-1}, \epsilon_{t,Y}) + 2Cov(X_{t-1}, \epsilon_{t,Y})
\end{align}
$$

Following a similar argument from the variance of $M$, we get the covariances here to be zero:

$$
\begin{align}
Var(Y_t) &= \eta_0^2Var(Y_{t-1}) + \eta_1^2 Var(M_{t-1}) + \eta_2^2 Var(X_{t-1}) + Var(\epsilon_{t,Y}) 0\\
Var(Y_t) - \eta_0^2Var(Y_{t-1}) &= \eta_1^2 \frac{\beta_1^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X + \sigma^2_M}{(1-\beta_0^2)} + \eta_2^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X + \sigma^2_Y\\
Var(Y_t)(1-\eta_0^2) &= \eta_1^2 \frac{\beta_1^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X + \sigma^2_M}{(1-\beta_0^2)} + \eta_2^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X + \sigma^2_Y\\
Var(Y_t) &= \frac{\eta_1^2 \frac{\beta_1^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X + \sigma^2_M}{(1-\beta_0^2)} + \eta_2^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X + \sigma^2_Y}{(1-\eta_0^2)} \\
Var(Y_t) &= \left( \frac{\eta^2_1\beta_1^2}{(1-\alpha^2)(1-\beta_0^2)(1-\eta_0^2)} + \frac{\eta^2_2}{(1-\alpha^2)(1-\eta_0^2)} \right) \sigma^2_X + \frac{\eta_1^2}{(1-\beta_0^2)(1-\eta_0^2)} \sigma^2_M + \frac{\sigma^2_Y}{(1-\eta_0^2)} \\
&= \left( \frac{\eta^2_1\beta_1^2 + \eta_2^2(1-\beta_0^2)}{(1-\alpha^2)(1-\beta_0^2)(1-\eta_0^2)}  \right) \sigma^2_X + \frac{\eta_1^2}{(1-\beta_0^2)(1-\eta_0^2)} \sigma^2_M + \frac{\sigma^2_Y}{(1-\eta_0^2)}
\end{align}
$$

From which we get an initial value for $Y$ of $Y_0 \sim N \left( 0 , \left( \eta_1^2 \frac{\beta_1^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X + \sigma^2_M}{(1-\beta_0^2)} + \eta_2^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X + \sigma^2_Y \right) /(1-\eta_0^2) \right)$.

We note that the values of $Y$ are dependent on two previous values of $X$, and so we also need an initial value for $Y_1$. Given the initial value for $Y$ above, then the value of $Y_1$ is $Y_1 \sim N \left( \eta_0 Var(Y_0) + \eta_1 Var(M_0) +\eta_2 Var(X_0), \left( \eta_1^2 \frac{\beta_1^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X + \sigma^2_M}{(1-\beta_0^2)} + \eta_2^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X + \sigma^2_Y \right) /(1-\eta_0^2) \right)$

### Covariances between X, M, and Y

It is a reasonable assumption to make that the covariance between the variables remains constant. We also assume that $Cov(\sigma_X^2,\sigma^2_M) = 0$, $Cov(\sigma_X^2,\sigma^2_Y) = 0$, and $Cov(\sigma_M^2,\sigma^2_Y) = 0$, while $Cov(\sigma_X^2,\sigma^2_X) = \sigma^2_X$ We also note that the covariance that we are interested in is

#### Cov(X,M)

For the covariance between $X$ and $M$, we get:

$$
\begin{align}
Cov(X,M) &= Cov(X_0,M_0) \\
&= Cov \left( \sigma^2_X \left( \frac{1}{1-\alpha^2} \right), \frac{\beta_1^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X + \sigma^2_M}{(1-\beta_0^2)} \right) \\
&= Cov \left( \sigma^2_X \left( \frac{1}{1-\alpha^2} \right), \frac{\beta_1^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X}{(1-\beta_0^2)} \right) \\
&= \frac{\beta_1^2}{(1-\alpha^2)^2(1-\beta_0^2)} \sigma^2_X
\end{align}
$$

#### Cov(X,Y)

Assuming constant covariances then we also find the covariances $Cov(X_t,Y_t)$, and $Cov(M_t,Y_t)$ to be:

$$
\begin{align}
Cov(X,Y) &= Cov \left(\left( \frac{1}{1-\alpha^2} \right)  \sigma^2_X  , \left( \frac{\eta^2_1\beta_1^2}{(1-\alpha^2)(1-\beta_0^2)(1-\eta_0^2)} + \frac{\eta^2_2}{(1-\alpha^2)(1-\eta_0^2)} \right) \sigma^2_X + \frac{\eta_1^2}{(1-\beta_0^2)(1-\eta_0^2)} \sigma^2_M + \frac{\sigma^2_Y}{(1-\eta_0^2)} \right) \\
&= \left( \frac{1}{1-\alpha^2} \right) \left( \frac{\eta^2_1\beta_1^2}{(1-\alpha^2)(1-\beta_0^2)(1-\eta_0^2)} + \frac{\eta^2_2}{(1-\alpha^2)(1-\eta_0^2)} \right) Cov(\sigma^2_X) \\
&= \left( \frac{1}{1-\alpha^2} \right) \left( \frac{\eta^2_1\beta_1^2}{(1-\alpha^2)(1-\beta_0^2)(1-\eta_0^2)} + \frac{\eta^2_2}{(1-\alpha^2)(1-\eta_0^2)} \right) \sigma^2_X \\
&= \frac{\eta_1^2\beta_1^2 + (1-\beta_0^2)\eta_2^2}{\left( 1-\alpha^2 \right) ^2(1-\beta_0^2) \left(1-\eta_0^2\right)} \sigma^2_X
\end{align}
$$

#### Cov(M,Y)

Following similar methods we find the covariance between $M$ and $Y$.

$$
\begin{align}
Cov(M,Y) &= Cov \left( \frac{\beta_1^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X}{(1-\beta_0^2)} + \frac{\sigma^2_M}{(1-\beta_0^2)(1-\eta_0^2)} , \left( \frac{\eta^2_1\beta_1^2 + \eta_2^2(1-\beta_0^2)}{(1-\alpha^2)(1-\beta_0^2)(1-\eta_0^2)}  \right) \sigma^2_X + \frac{\eta_1^2}{(1-\beta_0^2)(1-\eta_0^2)} \sigma^2_M + \frac{\sigma^2_Y}{(1-\eta_0^2)} \right) \\
&= Cov \left(  \frac{1}{(1-\beta_0^2)}\sigma^2_M, \frac{\eta_1^2}{(1-\beta_0^2)(1-\eta_0^2)} \sigma^2_M \right) + Cov \left( \frac{\beta_1^2 \left( \frac{1}{1-\alpha^2} \right) }{(1-\beta_0^2)} \sigma^2_X, \left( \frac{\eta^2_1\beta_1^2 + \eta_2^2(1-\beta_0^2)}{(1-\alpha^2)(1-\beta_0^2)(1-\eta_0^2)}  \right) \sigma^2_X \right)\\
&= \frac{1}{(1-\beta_0^2)} \cdot \frac{\eta_1^2}{(1-\beta_0^2)(1-\eta_0^2)} Cov(\sigma^2_M, \sigma^2_M) + \frac{\beta_1^2 }{(1-\beta_0^2)(1-\alpha^2)} \cdot \left( \left( \frac{\eta^2_1\beta_1^2 + \eta_2^2(1-\beta_0^2)}{(1-\alpha^2)(1-\beta_0^2)(1-\eta_0^2)}  \right)\right) Cov(\sigma^2_X,\sigma^2_X) \\
&= \frac{\eta_1^2}{\left( 1-\beta_0^2 \right)^2(1-\eta_0^2)} \sigma^2_M + \left(  \frac{\beta_1^4 \eta_1^2 + \beta_1^2\eta_2^2(1-\beta_0^2)}{\left(1-\alpha^2\right)^2 \left(1-\beta_0^2\right)^2 (1-\eta_0^2)}  \right) \sigma^2_X
\end{align}
$$

#### Estimation of Covariances

We note that these covariances rely on the values: $\sigma^2_X$, $\sigma^2_M$, and $\sigma^2_Y$. However we can estimate these values if we know the variance of the variables:

$$
\begin{align}
Var(X) &= \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X \\
\implies\sigma^2_X &= Var(X) (1-\alpha^2)
\end{align}
$$

Similarly for $\sigma^2_M$:

$$
\begin{align}
Var(M) &= \frac{\beta_1^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X + \sigma^2_M}{(1-\beta_0^2)} \\
\implies \sigma^2_M &= Var(M)(1-\beta_0^2) - \beta_1^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X
\end{align}
$$

And:

$$
\begin{align}
Var(Y) &= \left( \frac{\eta^2_1\beta_1^2 + \eta_2^2(1-\beta_0^2)}{(1-\alpha^2)(1-\beta_0^2)(1-\eta_0^2)}  \right) \sigma^2_X + \frac{\eta_1^2}{(1-\beta_0^2)(1-\eta_0^2)} \sigma^2_M + \frac{\sigma^2_Y}{(1-\eta_0^2)} \\
\implies \sigma^2_Y &= \left( Var(Y) - \left( \frac{\eta^2_1\beta_1^2 + \eta_2^2(1-\beta_0^2)}{(1-\alpha^2)(1-\beta_0^2)(1-\eta_0^2)}  \right) \sigma^2_X + \frac{\eta_1^2}{(1-\beta_0^2)(1-\eta_0^2)} \sigma^2_M \right) (1-\eta_0^2)
\end{align}
$$

### Autocovariances

#### ACF of X

We note that $X_t$ follows that AR(1) process and so it has the autocovariance $\gamma_x(h)$:

$$
\gamma_x(h) = \sigma^2_X \frac{\alpha^{|h|}}{(1-\alpha^2)}
$$

#### ACF of M

Now looking for the autocovariance of $M_t$ we start with $\gamma_m(0)$ which is the variance of $M_t$

$$
\gamma_m(0) = \frac{\beta_1^2 \left( \frac{1}{1-\alpha^2} \right) \sigma^2_X + \sigma^2_M}{(1-\beta_0^2)}
$$

Then $\gamma_m(1)$ is:

$$
\begin{align}
\gamma_m(1) &= Cov(M_t,M_{t-1}) \\
&= Cov(\beta_0 M_{t-1} + \beta_1 X_{t-1} + \epsilon_{M,t} ,M_{t-1}) \\
&= \beta_0Cov(M_{t-1}, M_{t-1}) + \beta_1 Cov(X_{t-1}, M_{t-1}) \\
&= \beta_0 \gamma_m(0) + \beta_1 Cov(X_{t-1}, M_{t-1}) 
\end{align}
$$

We also find the $\gamma_m(2)$: $$
\begin{align}
\gamma_m(2) &= Cov(M_t, M_{t-2}) \\
&= \beta_0Cov(M_{t-1}, M_{t-2}) + \beta_1 Cov(X_{t-1}, M_{t-2})\\
&= \beta_0 \gamma_m(1) + \beta_1 \alpha Cov(X_{t-2}, M_{t-2})
\end{align}
$$

And $\gamma_m(3)$:

$$
\begin{align}
\gamma_m(3) &= Cov(M_t, M_{t-3}) \\
&= \beta_0 \gamma_m(2) + \beta_1 Cov(X_{t-1}, M_{t-3}) \\
&= \beta_0 \gamma_m(2) + \beta_1 \alpha^2 Cov(X_{t-3}, M_{t-3})
\end{align}
$$

We now have enough observations to generalize this to the $\gamma_m(h)$ case:

$$
\begin{align}
\gamma_m(h) &= \beta_0^h \gamma_m(0) +  \sum^h_{i=1} \alpha^{i-1} \beta_1 Cov(X_{t-i}, M_{t-i})
\end{align}
$$

This becomes:

$$
\begin{align}
\gamma_m(h) &= \beta_0^h \gamma_m(0) + \beta_1 \left( \sum^h_{i=0} \alpha^i \right) Cov(X, M) \\
&= \beta_0^h \gamma_m(0) + \beta_1  \frac{1}{1-\alpha} \frac{\beta_1^2}{(1-\alpha^2)^2(1-\beta_0^2)} \sigma^2_X \\
&= \beta_0^h \gamma_m(0) \frac{\beta_1^3 \sigma^2_X}{(1-\alpha^2)^3(1-\beta_0^2)}
\end{align}
$$

#### ACF of Y

Using a similar approach to the one for the autocovariance of $M_t$ we start with $\gamma_y(0)$:

$$
\gamma_y(0) =  \left( \frac{\eta^2_1\beta_1^2 + \eta_2^2(1-\beta_0^2)}{(1-\alpha^2)(1-\beta_0^2)(1-\eta_0^2)}  \right) \sigma^2_X + \frac{\eta_1^2}{(1-\beta_0^2)(1-\eta_0^2)} \sigma^2_M + \frac{\sigma^2_Y}{(1-\eta_0^2)}
$$

We also find $\gamma_y(1)$:

$$
\begin{align}
\gamma_y(1) &= \eta_0 Cov(Y_{t-1},Y_{t-1}) + \eta_1 Cov(M_{t-1}, Y_{t-1}) + \eta_2 Cov(X_{t-1}, Y_{t-1}) \\
&= \eta_0 \gamma_y(0) + \eta_1 Cov(M_{t-1}, Y_{t-1}) + \eta_2 Cov(X_{t-1}, Y_{t-1})
\end{align}
$$

And $\gamma_y(2)$:

$$
\begin{align}
\gamma_y(2) &= \eta_0 Cov(Y_{t-1},Y_{t-2}) + \eta_1 Cov(M_{t-1}, Y_{t-2}) + \eta_2 Cov(X_{t-1}, Y_{t-2}) \\
&= \eta_0 \gamma_y(1) + \eta_1 \beta_0 Cov(M_{t-2},Y_{t-2}) + (\beta_1 \eta_1 + \eta_2 \alpha) Cov(X_{t-2},Y_{t-2})
\end{align}
$$

And $\gamma_y(3)$: $$
\begin{align}
\gamma_y(3) &= \eta_0\gamma_y(2) + \eta_1 Cov(M_{t-1}, Y_{t-3}) + \eta_2 Cov(X_{t-1}, Y_{t-3}) \\
&= \eta_0\gamma_y(2) + \eta_1(\beta_0Cov(M_{t-2},Y_{t-3}) + \beta_1 Cov(X_{t-2},Y_{t-3})) + \eta_2 \alpha^2 Cov(X_{t-3}, Y_{t-3}) \\
&= \eta_0 \gamma_y(2) + \eta_1\beta_0^2 Cov(M_{t-3}, Y_{t-3}) + (\eta_1\beta_1[\beta_0+\alpha]+\eta_2\alpha^2) Cov(X_{t-3}, Y_{t-3})
\end{align}
$$

We are beginning to see a pattern emerge, which one more iteration will make much more obvious:

$$
\begin{align}
\gamma_y(4) &= \eta_0 \gamma_y(3) + \eta_1 Cov(M_{t-1}, Y_{t-4}) + \eta_2 Cov(X_{t-1}, Y_{t-4}) \\
&= \eta_0 \gamma_y(3) + \eta_1 \beta_0^3 Cov(M_{t-4}, Y_{t-4}) + (\eta_1\beta_1[\beta_0^2+\beta_0\alpha+\alpha^2]+\eta_2\alpha^3)Cov(X_{t-4},Y_{t-4})
\end{align}
$$

From this we can get the generalization:

$$
\begin{align}
\gamma_y(h) &= \eta_0\gamma_y(h-1) + \eta_1\beta_0^hCov(M_{t-h},Y_{t-h}) + \left( [\beta_0^{h-2} + \beta_0^{h-3}\alpha + \cdots + \beta_0\alpha^{h-3} + \alpha^{h-2}] + \eta_2 \alpha^{h-1} \right) Cov(X_{t-h}, Y_{t-h}) \\
\gamma_y(h) &= \eta_0\gamma_y(h-1) + \eta_1\beta_0^hCov(M,Y) + \left( \frac{\beta_0^{h-1}-\alpha^{h-1}}{\beta_0 - \alpha} + \eta_2 \alpha^{h-1} \right) Cov(X, Y)
\end{align}
$$

### Interpretation of direct effect

We cannot use the same method we did for simple mediation since substituting in the values for $X_{t-1}$ and $M_{t-1}$ gets us:

$$
\begin{align}
Y_t &= \eta_0(\eta_0 Y_{t-2} + \eta_1 M_{t-2} + \eta_2 X_{t-2} + \epsilon_{t-1,Y}) + \eta_1(\beta_0 M_{t-2} +\beta_1 X_{t-2} +\epsilon_{t-1,M}) + \eta_2 (\alpha X_{t-2} + \epsilon_{t-1,X}) + \epsilon_{t,Y} \\
&= \eta_0^2 Y_{t-2} + (\eta_0 + \beta_0) \eta_1 M_{t-2} + (\eta_0 \eta_2 + \eta_1 \beta_1 + \eta_2 \alpha)X_{t-2} + (\eta_0 \epsilon_{t-1,Y} + \eta_1 \epsilon_{t-1,M} + \eta_2 \epsilon_{t-1,X} + \epsilon_{t,Y})
\end{align}
$$

From which we note that there are implicit $X_{t-3}$'s in $Y_{t-2}$ and $M_{t-2}$. Although this does allow us to make a statement about how $Y_{t}$ is related to $X_{t-2}$. The direct effect of $X_{t-2}$ on $Y_t$ is $\eta_0 \eta_2 + \eta_2 \alpha$, and the indirect effect of $X_{t-2}$ on $Y_t$ is $\eta_1 \beta_1$.

A natural question at this point is why not use $X_{t-3}$ since $X_{t-2}$, $M_{t-2}$, and $Y_{t-2}$ depend on $X_{t-3}$? There is no need to go this far back since it will not involve any new paths to get to $Y_t$, all of the inputs to $Y_t$ are captured when we consider $X_{t-2}$, and so we do not need to go back further. We also note that if we only look at the impact of $X_{t-1}$ on $Y_t$, then we will be missing the impact of $M_{t-1}$ and $Y_{t-1}$ since neither of them depend on $X_{t-1}$ and would not be captured if we were only examining the impact of $X_{t-1}$ on $Y_t$.

## Relationships

We note that there is no completely direct effect from $X_{t-2}$ to $Y_t$, in the sense of an immediate connection between the variables. The previous labelled direct effect is the effect of $X_{t-2}$ on $Y_t$ if $M_t$ is not present, and the indirect effect is the effect of $X_{t-2}$ on $Y_t$ when mediated through $M_{t-1}$. However both of the paths that the direct effect takes are themselves mediated through $X$ and $Y$. We consider the first option, $X_{t-2}$ mediated through itself $X_{t-1}$ to be the self-mediated direct effect, this is given by $\eta_2 \alpha$. There is also the second option of $X_{t-2}$ mediated through $Y_{t-1}$ which is the cross-mediated direct effect and is given by $\eta_0 \eta_2$.

We can summarize the total effect of $X_{t-2}$ on $Y_t$ in the following chart:

| Total Effect | $\eta_2 \alpha + \eta_0 \eta_2 + \eta_1 \beta_1$ |
|----|----|
| self-mediated direct effect | $\eta_2 \alpha$ |
| cross-mediated direct effect | $\eta_0 \eta_2$ |
| Indirect effect/mediated effect | $\eta_1 \beta_1$ |

## Further exploration of ARX(p) models

While ARX(1) models are a natural starting point, a next step is to determine what is the effect of a larger degrees of AR models?

For notational clarity we define ARX($p,i$) models, $p$ is the degree of the AR component, and $i$ is the number of terms of the first exogenous process that are included, if there is a second process then I will be referring to it as ARX($p,i,j$) and so on.

I will not be exploring much on this topic in this thesis, but it is an interesting area for future study.

## Simulation of theoretical results

## Conclusion of ARX Mediation

The derivations in this section explain mediation for ARX(1) models, including their theoretical variances and covariances.
