---
title: "Mediation_Details"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mediation_Details}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SimpleMediation)
```

This package is code for my thesis on mediation in a time series context. This vignette details the theory behind mediation and the functions that I am writing for the thesis. This vignette will help link the theory with the specific functions that I am writing.

# Basic mediation functions

The functions `run_mediation`, and `simple_mediation` are both for normal mediation on continuous variables. `run_mediation` runs a simple mediation on three variables, and `simple_mediation` uses it to perform bootstrapping in a simple mediation context and gives an estimate for the variance and covariance of the mediation.

# Mediation for ARX models

We denote an ARX(1) model as:

$$
X_t = \phi X_{t-1} + \psi Z_{t-1} + \epsilon_t
$$

Where: $\epsilon_t \sim N(0,\sigma^2)$

We can examine the autocovariance of an ARX(1) model. If we use the above definition the ARX(1) model is: $X_t = \phi X_{t-1} + \psi Z_{t-1} + \epsilon_t$, and suppose that $X_0 \sim WN(0, \sigma^2_X(\frac{1}{1-\phi})+\sigma^2_Z(\frac{\psi^2}{1-\psi^2}))$ and we find the autocovariance:

$$
\begin{align}
\gamma (0) &= \sigma^2_X(\frac{1}{1-\phi})+\sigma^2_Z(\frac{\psi^2}{1-\psi^2}) \\
\gamma(1) &= Cov(X_{t+1},X_t) \\
&= Cov(\phi X_t + \psi Z_t + \epsilon_t, X_t) \\
&= \phi Cov(X_t, X_t) + \psi Cov(Z_t,X_t) + Cov(\epsilon_t, X_t) \\
&= \phi \gamma(0) + \psi Cov(Z_t,X_t)
\end{align}
$$

This calculation depends on the value of $Cov(Z_t,X_t)$:

$$
\begin{align}
Cov(Z_t,X_t) &= Cov(Z_t, \phi X_{t-1} + \psi Z_{t-1} + \epsilon_{z,t}) \\
&= \phi Cov(Z_t,X_{t-1}) + \psi Cov(Z_t,Z_{t-1}) + Cov(Z_t,\epsilon_{z,t}) \\
&= \phi Cov(Z_t, X_{t-1}) + \psi\gamma_z(1) \\
&= \phi(\phi Cov(Z_t,X_{t-2}) + \psi \gamma_z (2)) + \psi \gamma_z (1)
\end{align}
$$

## ARX in Mediation

For mediation we define $X$ to be AR(1) and $M$ and $Y$ to be ARX(1):

$$
\begin{align}
X_t &= \alpha X_{t-1} + \epsilon_{t,X}\\
M_t &= \beta_0 M_{t-1} +\beta_1 X_{t-1} +\epsilon_{t,M} \\
Y_t &= \eta_0 Y_{t-1} + \eta_1 M_{t-1} + \eta_2 X_{t-1} + \epsilon_{t,Y}
\end{align}
$$

In this context, since $X_t$, $M_t$, and $Y_t$ are all AR based models we also define $\alpha$, $\beta_0$, and $\eta_0$ to be bounded between -1 and 1:

$$
\alpha,\beta_0,\eta_0 \in (-1,1)
$$

It is reasonable for us to assume that each of these process is stationary. While many time series are not actually stationary, there are a wide variety of techniques that transform a non-stationary time series into a stationary one, it is reasonable for us to assume that when using this process the starting time series can be turned into stationary ones.

We also define the errors from these models to be:

$$
\epsilon_{t,X} \sim N(0,\sigma^2_X)\\
\epsilon_{t,M} \sim N(0,\sigma^2_M)\\
\epsilon_{t,Y} \sim N(0,\sigma^2_Y)
$$

Here we are also assuming that the errors are uncorrelated with each other.

### Initial Values

For an AR(1) model the initial point $X_0$ follows a distribution with mean 0, and variance $\sigma^2/(1-\phi^2)$. In the mediation context we will need initial values for $X$, $M$, and $Y$.

Since $X$ is an AR(1) model, we define the initial value to be:

$$
X_0\sim N \left( 0,\frac{\sigma^2_X}{(1-\phi^2)} \right)
$$

We also need an initial value for $M$, which is an ARX(1) process. The mean of the initial value is still 0, but the variance is now also dependent on $X$.

### Options for Var(M)

$$
\begin{align}
Var(M_t) &= Var(\beta_0 M_{t-1} + \beta_1 X_{t-1} + \epsilon_{t,M}) \\
&= Var( \epsilon_{t,M} + \beta_0 \epsilon_{t-1,M} + \beta_0^2 \epsilon_{t-2,M} + \cdots +
\beta_1 \epsilon_{t-1,X} + \beta_1^2 \epsilon_{t-2,X} + \cdots ) \\
&= \sigma^2_M (1 + \beta_0 + \beta_0^2+ \cdots) + \sigma^2_X (\beta_1 + \beta_1^2 + \cdots) \\
&= \sigma^2_M \left( \frac{1}{1-\beta_0^2} \right) + \sigma^2_X \left( \frac{\beta_1^2}{1-\beta_1^2} \right)
\end{align}
$$

OR

$$
\begin{align}
Var(M_t) &= Var(\beta_0 M_{t-1} + \beta_1 X_{t-1} + \epsilon_{t,M}) \\
&= Var(\epsilon_{t,M} + \beta_0 (\beta_0 M_{t-2} + \beta_1 X_{t-2} + \epsilon_{t-1,M}) + \beta_1 X_{t-1}) \\
&= Var( \epsilon_{t,M} + \beta_0 \epsilon_{t-1,M} + \beta_0^2 \epsilon_{t-2,M} + \cdots +
\beta_1 ( X_{t-1} + \beta_0 X_{t-2} + \cdots ) ) \\
&= \sigma^2_M (1 + \beta_0 + \beta_0^2+ \cdots) + \frac{\sigma^2_X \beta_1}{1-\alpha^2} (1 + \beta_0 + \beta_0^2 + \cdots)  \\
&= \sigma^2_M \left( \frac{1}{1-\beta_0^2} \right) + \frac{\sigma^2_X \beta_1}{1-\alpha^2} \left( \frac{1}{1-\beta_0^2} \right)
\end{align}
$$

OR

$$
\begin{align}
Var(M_t) &= Var(\beta_0 M_{t-1} + \beta_1 X_{t-1} + \epsilon_{t,M}) \\
&= Var(\beta_0 M_{t-1}) + Var(\beta_1 X_{t-1}) +2 Cov(X_{t-1}, M_{t-1}) + \sigma^2_M \\
&= \beta_0^2Var(M_{t-1}) + \beta^2_1 Var(X_{t-1}) + 2 Cov(X_{t-1}, M_{t-1}) + \sigma^2_M \\
Var(M_t) - \beta_0^2 Var(M_{t-1}) &= \beta_1^2 \frac{1}{(1-\alpha^2)} \sigma^2_X + 2Cov(X_{t-1},M_{t-1}) + \sigma^2_M \\
(1-\beta_0^2) Var(M_t) &= \beta_1^2 \frac{1}{(1-\alpha^2)} \sigma^2_X + 2Cov(X_{t-1},M_{t-1}) + \sigma^2_M \\
Var(M_t) &= \frac{\beta_1^2 \frac{1}{(1-\alpha^2)} \sigma^2_X + 2Cov(X_{t-1},M_{t-1}) + \sigma^2_M}{(1-\beta_0^2)}
\end{align}
$$

So we get an initial value for $M$ of $M_0 \sim N \left( 0, \sigma^2_M \left( \frac{1}{1-\beta_0^2} \right) + \sigma^2_X \left( \frac{\beta_1^2}{1-\beta_1^2} \right) \right)$.

Following a similar argument we get an initial value for $Y$ of $Y_0 \sim N \left( 0 , \sigma^2_Y \left( \frac{1}{1-\eta_0^2} \right) + \sigma^2_M \left( \frac{\eta_1^2}{1-\eta_1^2} \right) + \sigma^2_X \left( \frac{\eta_2^2}{1-\eta_2^2} \right) \right)$.

We note that the values of $Y$ are dependent on two previous values of $X$, and so we also need an initial value for $Y_1$. If we say that the initial value of $Y_0$ is $Y_0 \sim N \left( 0 , \sigma^2_Y \left( \frac{1}{1-\eta_0^2} \right) + \sigma^2_M \left( \frac{\eta_1^2}{1-\eta_1^2} \right) + \sigma^2_X \left( \frac{\eta_2^2}{1-\eta_2^2} \right) \right)$. Then the value of $Y_1$ is $Y_1 \sim N \left( \eta_0 Var(Y_0) + \eta_1 Var(M_0) +\eta_2 Var(X_0), \sigma^2_Y \left( \frac{1}{1-\eta_0^2} \right) + \sigma^2_M \left( \frac{\eta_1^2}{1-\eta_1^2} \right) + \sigma^2_X \left( \frac{\eta_2^2}{1-\eta_2^2} \right) \right)$

### Autocovariances

#### ACF of X

We note that $X_t$ follows that AR(1) process and so it has the autocovariance $\gamma_x(h)$:

$$
\gamma_x(h) = \sigma^2_X \frac{\phi^{|h|}}{(1-\phi^2)}
$$

#### ACF of M

Now looking for the autocovariance of $M_t$ we start with $\gamma_m(0)$ which is the variance of $M_t$ $$
\gamma_m(0) = \sigma^2_M \left( \frac{1}{1-\beta_0^2} \right) + \sigma^2_X \left( \frac{\beta_1^2}{1-\beta_1^2} \right)
$$

Then $\gamma_m(1)$ is:

$$
\begin{align}
\gamma_m(1) &= Cov(M_t,M_{t-1}) \\
&= Cov(\beta_0 M_{t-1} + \beta_1 X_{t-1} + \epsilon_{M,t} ,M_{t-1}) \\
&= \beta_0Cov(M_{t-1}, M_{t-1}) + \beta_1 Cov(X_{t-1}, M_{t-1}) \\
&= \beta_0 \gamma_m(0) + \beta_1 Cov(X_{t-1}, M_{t-1}) 
\end{align}
$$

We also find the $\gamma_m(2)$: $$
\begin{align}
\gamma_m(2) &= Cov(M_t, M_{t-2}) \\
&= \beta_0Cov(M_{t-1}, M_{t-2}) + \beta_1 Cov(X_{t-1}, M_{t-2})\\
&= \beta_0 \gamma_m(1) + \beta_1 \alpha Cov(X_{t-2}, M_{t-2})
\end{align}
$$

And $\gamma_m(3)$:

$$
\begin{align}
\gamma_m(3) &= Cov(M_t, M_{t-3}) \\
&= \beta_0 \gamma_m(2) + \beta_1 Cov(X_{t-1}, M_{t-3}) \\
&= \beta_0 \gamma_m(2) + \beta_1 \alpha^2 Cov(X_{t-3}, M_{t-3})
\end{align}
$$

We now have enough observations to generalize this to the $\gamma_m(h)$ case:

$$
\begin{align}
\gamma_m(h) &= \beta_0^h \gamma_m(0) +  \sum^h_{i=1} \alpha^{i-1} \beta_1 Cov(X_{t-i}, M_{t-i})
\end{align}
$$

If we assume that the covariance is constant, this becomes:

$$
\begin{align}
\gamma_m(h) &= \beta_0^h \gamma_m(0) + \beta_1 Cov(X_{t}, M_{t}) \sum^h_{i=0} \alpha^i\\
&= \beta_0^h \gamma_m(0) + \beta_1 Cov(X_{t}, M_{t}) \cdot \frac{1}{1-\alpha}
\end{align}
$$

This relies on knowing the covariance between $X_t$ and $M_t$.

#### ACF of Y

Using a similar approach to the one for the autocovariance of $M_t$ we start with $\gamma_y(0)$:

$$
\gamma_y(0) = \sigma^2_Y \left( \frac{1}{1-\eta_0^2} \right) + \sigma^2_M \left( \frac{\eta_1^2}{1-\eta_1^2} \right) + \sigma^2_X \left( \frac{\eta_2^2}{1-\eta_2^2} \right)
$$

We also find $\gamma_y(1)$:

$$
\begin{align}
\gamma_y(1) &= \eta_0 Cov(Y_{t-1},Y_{t-1}) + \eta_1 Cov(M_{t-1}, Y_{t-1}) + \eta_2 Cov(X_{t-1}, Y_{t-1}) \\
&= \eta_0 \gamma_y(0) + \eta_1 Cov(M_{t-1}, Y_{t-1}) + \eta_2 Cov(X_{t-1}, Y_{t-1})
\end{align}
$$

And $\gamma_y(2)$:

$$
\begin{align}
\gamma_y(2) &= \eta_0 Cov(Y_{t-1},Y_{t-2}) + \eta_1 Cov(M_{t-1}, Y_{t-2}) + \eta_2 Cov(X_{t-1}, Y_{t-2}) \\
&= \eta_0 \gamma_y(1) + \eta_1 \beta_0 Cov(M_{t-2},Y_{t-2}) + (\beta_1 \eta_1 + \eta_2 \alpha) Cov(X_{t-2},Y_{t-2})
\end{align}
$$

And $\gamma_y(3)$: $$
\begin{align}
\gamma_y(3) &= \eta_0\gamma_y(2) + \eta_1 Cov(M_{t-1}, Y_{t-3}) + \eta_2 Cov(X_{t-1}, Y_{t-3}) \\
&= \eta_0\gamma_y(2) + \eta_1(\beta_0Cov(M_{t-2},Y_{t-3}) + \beta_1 Cov(X_{t-2},Y_{t-3})) + \eta_2 \alpha^2 Cov(X_{t-3}, Y_{t-3}) \\
&= \eta_0 \gamma_y(2) + \eta_1\beta_0^2 Cov(M_{t-3}, Y_{t-3}) + (\eta_1\beta_1[\beta_0+\alpha]+\eta_2\alpha^2) Cov(X_{t-3}, Y_{t-3})
\end{align}
$$

We are beginning to see a pattern emerge, which one more iteration will make much more obvious:

$$
\begin{align}
\gamma_y(4) &= \eta_0 \gamma_y(3) + \eta_1 Cov(M_{t-1}, Y_{t-4}) + \eta_2 Cov(X_{t-1}, Y_{t-4}) \\
&= \eta_0 \gamma_y(3) + \eta_1 \beta_0^3 Cov(M_{t-4}, Y_{t-4}) + (\eta_1\beta_1[\beta_0^2+\beta_0\alpha+\alpha^2]+\eta_2\alpha^3)Cov(X_{t-4},Y_{t-4})
\end{align}
$$

From this we can get the generalization:

$$
\begin{align}
\gamma_y(h) &= \eta_0\gamma_y(h-1) + \eta_1\beta_0^hCov(M_{t-h},Y_{t-h}) + \left( [\beta_0^{h-2} + \beta_0^{h-3}\alpha + \cdots + \beta_0\alpha^{h-3} + \alpha^{h-2}] + \eta_2 \alpha^{h-1} \right) Cov(X_{t-h}, Y_{t-h}) \\
\gamma_y(h) &= \eta_0\gamma_y(h-1) + \eta_1\beta_0^hCov(M_{t-h},Y_{t-h}) + \left( \frac{\beta_0^{h-1}-\alpha^{h-1}}{\beta_0 - \alpha} + \eta_2 \alpha^{h-1} \right) Cov(X_{t-h}, Y_{t-h})
\end{align}
$$

### Covariances between X, M, and Y

We note that these values all directly rely on $Cov(X_t,M_t)$, $Cov(X_t,Y_t)$, and $Cov(M_t,Y_t)$.

There are two possibilities here. If we assume that all the covariances are the same then:

$$
\begin{align}
Cov(X_t,M_t) &= Cov(X_0,M_0) \\
&= Cov \left( \sigma^2_X \left( \frac{1}{1-\alpha^2} \right), \sigma^2_M \left( \frac{1}{1-\beta_0^2} \right) + \sigma^2_X \left( \frac{\beta_1^2}{1-\beta_1^2} \right) \right) \\
&= Cov \left( \sigma^2_X \left( \frac{1}{1-\alpha^2} \right), \sigma^2_X \left( \frac{\beta_1^2}{1-\beta_1^2} \right) \right) \\
&= \frac{\beta_1^2}{(1-\alpha^2)(1-\beta_1^2)} \sigma^2_X
\end{align}
$$

Since $Cov(\sigma^2_X,\sigma^2_M)=0$.

If the assumption of constant covariances does not hold then we have the following:

$$
\begin{align}
Cov(X_t,M_t) &= \alpha^t Cov(X_0,M_t) \\
&= \alpha^t (\beta_0Cov(X_0,M_{t-1}) + \beta_1 Cov(X_0,X_{t-1})) \\
&= \alpha^t\beta_0 Cov(X_0,M_0) + \beta_1 \sum_{i=1}^{t-1} \beta_0^i\gamma_x(t-i) \\
&= \alpha^t\beta_0 \left( \frac{\beta_1^2}{(1-\alpha^2)(1-\beta_1^2)} \sigma^2_X \right) + \beta_1 \sum_{i=1}^{t-1} \beta_0^i\gamma_x(t-i)
\end{align}
$$

Assuming constant covariances then we also find the covariances $Cov(X_t,Y_t)$, and $Cov(M_t,Y_t)$ to be:

$$
\begin{align}
Cov(X_t,Y_t) &= Cov \left( \sigma^2_X \left( \frac{1}{1-\alpha^2} \right), \sigma^2_Y \left( \frac{1}{1-\eta_0^2} \right) + \sigma^2_M \left( \frac{\eta_1^2}{1-\eta_1^2} \right) + \sigma^2_X \left( \frac{\eta_2^2}{1-\eta_2^2} \right) \right) \\
&= Cov \left( \sigma^2_X \left( \frac{1}{1-\alpha^2} \right),  \sigma^2_X \left( \frac{\eta_2^2}{1-\eta_2^2} \right) \right) \\
&= \frac{\eta_2^2}{(1-\alpha^2)(1-\eta_2^2)} \sigma^2_X
\end{align}
$$

and:

$$
\begin{align}
Cov(M_t,Y_t) &= Cov \left( \sigma^2_M \left( \frac{1}{1-\beta_0^2} \right) + \sigma^2_X \left( \frac{\beta_1^2}{1-\beta_1^2} \right) , \sigma^2_Y \left( \frac{1}{1-\eta_0^2} \right) + \sigma^2_M \left( \frac{\eta_1^2}{1-\eta_1^2} \right) + \sigma^2_X \left( \frac{\eta_2^2}{1-\eta_2^2} \right) \right) \\
&= Cov \left(  \sigma^2_M \left( \frac{1}{1-\beta_0^2} \right), \sigma^2_M \left( \frac{\eta_1^2}{1-\eta_1^2} \right) \right) + Cov \left( \sigma^2_X \left( \frac{\beta_1^2}{1-\beta_1^2} \right), \sigma^2_X \left( \frac{\eta_2^2}{1-\eta_2^2} \right) \right)\\
&= \left( \frac{\eta_1^2}{(1-\beta^2_0)(1-\eta_1^2)} \right) Cov(\sigma^2_M, \sigma^2_M) + \left( \frac{\beta_1^2\eta_2^2}{(1-\beta_1^2)(1-\eta^2_2)} \right) Cov(\sigma^2_X,\sigma^2_X) \\
&= \left( \frac{\eta_1^2}{(1-\beta^2_0)(1-\eta_1^2)} \right) \sigma^2_M + \left( \frac{\beta_1^2\eta_2^2}{(1-\beta_1^2)(1-\eta^2_2)} \right) \sigma^2_X
\end{align}
$$

### Interpretation of direct effect

We cannot use the same method we did for simple mediation since substituting in the values for $X_{t-1}$ and $M_{t-1}$ gets us:

$$
\begin{align}
Y_t &= \eta_0(\eta_0 Y_{t-2} + \eta_1 M_{t-2} + \eta_2 X_{t-2} + \epsilon_{t-1,Y}) + \eta_1(\beta_0 M_{t-2} +\beta_1 X_{t-2} +\epsilon_{t-1,M}) + \eta_2 (\alpha X_{t-2} + \epsilon_{t-1,X}) + \epsilon_{t,Y} \\
&= \eta_0^2 Y_{t-2} + (\eta_0 + \beta_0) \eta_1 M_{t-2} + (\eta_0 \eta_2 + \eta_1 \beta_1 + \eta_2 \alpha)X_{t-2} + (\eta_0 \epsilon_{t-1,Y} + \eta_1 \epsilon_{t-1,M} + \eta_2 \epsilon_{t-1,X} + \epsilon_{t,Y})
\end{align}
$$

From which we note that there are implicit $X_{t-3}$'s in $Y_{t-2}$ and $M_{t-2}$. Although this does allow us to make a statement about how $Y_{t}$ is related to $X_{t-2}$. The direct effect of $X_{t-2}$ on $Y_t$ is $\eta_0 \eta_2 + \eta_2 \alpha$, and the indirect effect of $X_{t-2}$ on $Y_t$ is $\eta_1 \beta_1$.

A natural question at this point is why not use $X_{t-3}$ since $X_{t-2}$, $M_{t-2}$, and $Y_{t-2}$ depend on $X_{t-3}$? There is no need to go this far back since it will not involve any new paths to get to $Y_t$, all of the inputs to $Y_t$ are captured when we consider $X_{t-2}$, and so we do not need to go back further. We also note that if we only look at the impact of $X_{t-1}$ on $Y_t$, then we will be missing the impact of $M_{t-1}$ and $Y_{t-1}$ since neither of them depend on $X_{t-1}$ and would not be captured if we were only examining the impact of $X_{t-1}$ on $Y_t$.

## Relationships

We note that there is no completely direct effect from $X_{t-2}$ to $Y_t$, in the sense of an immediate connection between the variables. The previous labelled direct effect is the effect of $X_{t-2}$ on $Y_t$ if $M_t$ is not present, and the indirect effect is the effect of $X_{t-2}$ on $Y_t$ when mediated through $M_{t-1}$. However both of the paths that the direct effect takes are themselves mediated through $X$ and $Y$. We consider the first option, $X_{t-2}$ mediated through itself $X_{t-1}$ to be the self-mediated direct effect, this is given by $\eta_2 \alpha$. There is also the second option of $X_{t-2}$ mediated through $Y_{t-1}$ which is the cross-mediated direct effect and is given by $\eta_0 \eta_2$.

We can summarize the total effect of $X_{t-2}$ on $Y_t$ in the following chart:

| Total Effect | $\eta_2 \alpha + \eta_0 \eta_2 + \eta_1 \beta_1$ |
|----|----|
| self-mediated direct effect | $\eta_2 \alpha$ |
| cross-mediated direct effect | $\eta_0 \eta_2$ |
| Indirect effect/mediated effect | $\eta_1 \beta_1$ |

## Further exploration of ARX(p) models

While ARX(1) models are a natural starting point, what is the effect of larger degrees of AR models?

For notational clarity we define ARX($p,i$) models, $p$ is the degree of the AR component, and $i$ is the number of terms of the first exogenous process that are included, if there is a second process then I will be referring to it as ARX($p,i,j$) and so on.
