---
title: "Thesis_Draft_1"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Thesis_Draft_1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SimpleMediation)
```

# Abstract

# Introduction

# Background/Lit review

## Simple Mediation

## General Non-linear approach

## General Spline stuff

## General time series stuff 

(needed since we're doing a bunch with ARX stuff)
Some background is needed on time series models, particularly auto-regressive (AR) and auto-regressive models with exogenous inputs (ARX). The 

Reference the book for the time series course I did

This should be a pretty brief section

## Current non-linear mediation stuff

There has been some recent work in the field of mediation analysis, 

A unique aspect of this work is that it assumes a time series structure to the data using ARX models, which includes more causalities than a more traditional mediation relationship. There has been some work done on continuous mediators and treatments, and some on the non-linearity in mediation.

# Method

## Spline stuff from proposal?

## ARX based method

Just copy and paste for this part!

## Prediction based method

Same as above, but add section for how the methods compare in simulations!

## Future Work

Mostly in the assumed structure of the data (have some parts written about this already)

# Conclusion